### 2023-10-31 chapters/background.tex
Update background with IWAE

\todo[inline]{Come up with a suitable example explaining the intuition behind information entropy.}

@@

\section{Out-of-distribution detection} \label{sec:out-of-distribution-detection}
\todo[inline]{Maybe this section should come after the VAE background?}
% 
% OUTLINE: C.f. https://openaccess.thecvf.com/content/CVPR2023W/VAND/papers/Graham_Denoising_Diffusion_Models_for_Out-of-Distribution_Detection_CVPRW_2023_paper.pdf
% - Formalize the problem. https://arxiv.org/pdf/2302.10326.pdf
%   - Assume we have a dataset $\mathcal{D}_{\text{train}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from $\mathcal{D}_{\text{train}}$ or not. Given a test point $\xb$, the model outputs a score $s(\xb)$ that indicates whether $\xb$ is in-distribution or out-of-distribution. 
%   - In the context of neural networks, this problem dates back to at least 1994 \parencite{bishop_novelty_1994}. 
%
% - Categorisation into supervised and unsupervised. 
%   - Supervised: Whether p(y|x) should be trusted for a given x. 
%   - Unsupervised: Whether x should be trusted and used at all. 
%
In this section we introduce out-of-distribution (OOD) detection and connect it to the concept of uncertainty. 
Since the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic} are relatively short, we also provide a concise review of the literature on out-of-distribution detection.

\todo[inline]{Briefly hold the readers hand at actualize the importance of OODD.}

@@

% No OOD, feature
The final category of supervised methods derive the score from a latent representation of the input. 
Some methods define OOD classes. 
\textcite{huang_mos_2021} group the classes of the target variable $y$ and defines an OOD class for each group. Each training example is then the correct target for one group and an OOD example for all other groups; a kind of hierarchical version of the OOD class of simpler baselines based on noise augmentation \parencite{ren_likelihood_2019}. 
\todo[inline]{Does the previous sentence belong here or in a different paragraph? It is not really using the latent space.}
Similarly, to represent a virtual OOD class, \textcite{wang_vim_2022} generate an additional logit by first computing the residual of the input's latent space representation against a principal feature space and then converting it to a valid logit by matching its mean over training samples to the average maximum logits.

@@

\subsection{Variational inference} \label{sec:variational-autoencoders}
To be able to model high-dimensional data, we want to use neural networks to represent the generative process. 
This leads to deep latent variable models which were pioneered by \textcite{kingma_autoencoding_2014,rezende_stochastic_2014}. The authors propose to factor the joint distribution and parameterize it using a neural network with parameters $\thetab$ and let $p_\thetab(\xb,\zb) = p_\thetab(\xb|\zb)p(\zb)$ where $p(\zb)$ is the prior.

This choice makes both the marginal and posterior distributions intractable since they involve an integral over all possible values of the latent variable $\zb$. 
This precludes training with commonly used methods such as expectation maximization and exact maximum likelihood estimation and while MCMC methods are usually too computationally expensive for large neural networks, variational inference provides a scalable alternative \parencite{jordan_introduction_1999}.

@@

\paragraph{Evidence lower bound}
By estimating the posterior distribution $p(\zb|\xb)$ with a variational approximation $q(\zb)$ we can derive a lower bound on the otherwise intractable marginal log-likelihood $\log p_\theta(\xb)$.
%
 \begin{align} \label{eq:variational-autoencoder-elbo}
    \log p_\thetab(\xb)
    &= \log \int p_\thetab(\xb,\zb),\textrm{d}\zb \nonumber \\
    &= \log \int q(\zb) \frac{p_\thetab(\xb,\zb)}{q(\zb)}\,\textrm{d}\zb \nonumber \\
    &\geq \int q(\zb) \log \left( \frac{p_\thetab(\xb,\zb)}{q(\zb)} \right) \,\textrm{d}\zb \nonumber \\
    &\geq \mathbb{E}_{q(\zb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q(\zb)} \right] \equiv \mathcal{L}_{}(\xb;q) \enspace .
    % &= \log \int p_\thetab(\xb|\zb)p(\zb),\textrm{d}\zb \nonumber \\
    % &= \log \int q(\zb) \frac{p_\thetab(\xb|\zb)p(\zb)}{q(\zb)}\,\textrm{d}\zb \nonumber \\
    % &\geq \int q(\zb) \log \left( \frac{p_\thetab(\xb|\zb)p(\zb)}{q(\zb)} \right) \,\textrm{d}\zb \nonumber \\
    % &\geq \mathbb{E}_{q(\zb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q(\zb)} \right] \equiv \mathcal{L}_{}(\xb;\thetab,\phib) \enspace .
\end{align}
%
Since we can evaluate the generative model $p_\thetab(\xb,\zb)$ and are free to define $q(\zb)$ as we please, this bound can be made tractable.

The usual approach in variational inference is to fit an approximate posterior $q(\zb)$ per data point. However, for large datasets, as is common in natural language processing, this can be too computationally expensive. 
% \moretodo[inline]{Glue this together..}
Instead, VAEs amortize the cost of variational inference by parameterizing the approximate posterior with a neural network that transforms any data point $\xb$ into the parameters of the corresponding conditional distribution, $q_\phib(\zb|\xb)$. This makes it possible to efficiently infer the approximate posterior distribution for any input $\xb$. 
With amortized variational inference, the ELBO can simply be written as,
% The approximate posterior can also be used to compute a lower bound on the marginal distribution $p_\thetab(\xb)$ via Jensen's inequality, the so-called evidence lower bound (ELBO), 
%
\begin{equation}
    \log p_\thetab(\xb) \geq \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \equiv \mathcal{L}_{}(\xb;\thetab,\phib) \enspace .
\end{equation}
%
Although analytically evaluating the ELBO is intractable due to the expectation, it can be approximated by sampling and then used to train VAEs. We will return to this after examining the ELBO in more detail.

\paragraph{KL-divergence to prior}
Since the joint factorizes in VAEs, we can also write the ELBO in \cref{eq:variational-autoencoder-elbo} as,
%
\begin{align} \label{eq:variational-autoencoder-elbo-factorized}
    \mathcal{L}_{}(\xb;\thetab,\phib)
    % &\equiv \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
    &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right] + \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
    &= 
    \underset{\text{reconstruction loss}}{\underbrace{
        \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]
    }}
     - 
    \underset{\text{KL-divergence to prior}}{
        \underbrace{\vphantom{\mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]}{D_\text{KL}\left( q_\phib(\zb|\xb) \parallel p(\zb) \right)}
    }} \enspace .
\end{align}
%
This reveals a commonly used interpretation of the ELBO: 
The expected log-likelihood of the generative model $\mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]$ acts as a reconstruction loss that encourages the model to make $q_\phib(\zb|\xb)$ as peaky as possible since mapping many $\zb$ to a single $\xb$ is hard. 
The negative KL-divergence between the approximate posterior $q_\phib(\zb|\xb)$ and the prior $p(\zb)$ acts as a regularizer that forces the approximate posterior to be within the support of the prior (reverse KL-divergence). This helps prevent overfitting but importantly also enables ancestral sampling from the prior to generate high quality data.

@@

We can analyze the inference gap further \cite{cremer_inference_2018}. 
By amortizing the variational inference we introduce a gap compared to the alternative of fitting a posterior per data point, $q^{*}_{\xb}(\zb)$. This is referred to as the amortization gap and can be reduced by using large capacity amortized posteriors. 
In case the true posterior does not belong to the model class of the variational family of posteriors that we fit, the inference gap cannot be reduced to zero and the remaining gap is referred to as the approximation gap. 
If the inference gap is closed, the maximum of the ELBO corresponds to the maximum of the maringa
Finally, the generative model

\moretodo[inline]{Finish this with the parsimony gap}

A final error is introduced  the maximally attainable likelihood over all the group $G$ of possible models, $\mathcal{l}(G)$

It is important to also note that in fitting a VAE, the approximation gap is not the only source of error.

\moretodo[inline]{Maybe add a small graphic illustrating the gaps.}

@@

\paragraph{Sampling}
Since we have freedom to select the approximate posterior, we can choose it such that it is easy to sample and then use that property to form a Monte Carlo estimate of the ELBO,
% 
\begin{equation} \label{eq:variational-autoencoder-elbo-sampling-expanded}
    \mathcal{L}_{}(\xb;\thetab,\phib)
    \approx \frac{1}{K} \sum_{k=1}^K \log p_\thetab(\xb|\zb_k) + \log p(\zb_k) - \log q_\phib(\zb_k|\xb) \enspace , 
\end{equation}
%
where $\zb_k \sim q_\phib(\zb|\xb)$ are samples from the approximate posterior. 
For some choices of prior and approximate posterior, such as Gaussians, the KL-divergence term of the ELBO can be computed analytically. 
This Monte Carlo estimator can be written as,
%
\begin{equation} \label{eq:variational-autoencoder-elbo-sampling-KL}
    \mathcal{L}_{}(\xb;\thetab,\phib)
    \approx \frac{1}{K} \sum_{k=1}^K \log p_\thetab(\xb|\zb_k) - D_\text{KL}\left( q_\phib(\zb|\xb) \,||\, p(\zb) \right) \enspace .
\end{equation}
%
To reduce the variance of the estimate, \textcite{burda_importance_2016} propose to use importance sampling to reweigh the samples. 
Compared to the estimator in \cref{eq:variational-autoencoder-elbo-sampling-expanded,eq:variational-autoencoder-elbo-sampling-KL} which

\moretodo[inline]{Derive and introduce the IWAE bound.}

\paragraph{Reparameterization} 
% We want to use maximum likelihood estimation to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$. 
% Gradient-based optimization works well for neural networks, but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
Computing the gradient of the ELBO with respect to $\thetab$ is straightforward,
%
\begin{equation}
    % \nabla_\thetab \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] 
    \nabla_\thetab \mathcal{L}_{}(\xb;\thetab,\phib)
    = \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \nabla_\thetab \log p_\thetab(\xb|\zb) \right] \enspace .
\end{equation}
%
However, the gradient with respect to the $\phib$ is more challenging. Since $\phib$ occurs in the expectation itself, we may not move the gradient inside the expectation. 
However, by reparameterizing the latent variable $\zb$ as a deterministic function of the input $\xb$ and a random variable $\epsilonb\sim p(\epsilonb)$, $\zb = g_\phib(\xb,\epsilonb)$, we can write the gradient as a path-wise derivative \parencite{kingma_autoencoding_2014,rezende_stochastic_2014,mohamed_monte_2019},
%
\begin{equation}
    % \nabla_\phib \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] 
    \nabla_\phib \mathcal{L}_{}(\xb;\thetab,\phib)
    = \mathbb{E}_{p(\epsilonb)} \left[ \nabla_\phib \log \frac{p_\thetab\left(\xb|g_\phib(\xb,\epsilonb)\right)p\left(g_\phib(\xb,\epsilonb)\right)}{q_\phib\left(g_\phib(\xb,\epsilonb)|\xb\right)} \right] \enspace . \\
\end{equation}
%
The most important instantiation of this reparameterization is the diagonal Gaussian case. 
By choosing $q_\phib(\zb|\xb) = \mathcal{N}(\zb; \mub_\phib(\xb), \mathbf{I}\sigmab_\phib(\xb))$ it is possible to sample differentiably via the reparameterization trick by setting $g_\phib(\xb,\epsilonb) = \mu_\phib(\xb) + \sigma_\phib(\xb) \odot \epsilonb$ with $\epsilonb\sim\mathcal{N}(0,\mathbf{I})$. 
By assuming the covariance matrix is diagonal, this becomes a mean-field approximation which imposes independence between the dimensions of $\zb$.

@@

\subsection{Hierarchical models}\label{sec:hierarchical-vae-background}
The mean-field assumption imposed by $\zb \sim\mathcal{N}(\mub,\mathbf{I}\sigmab^2)$ can often be more restrictive for model expressiveness than we would like. 
While learning a full covariance is a simple solution, it is not always sufficiently computationally efficient and, in any case, enables learning only linear covariance between elements of $\zb$. 
The drawbacks of mean-field approximation and linear covariance have lead to research into learning hierarchies of several non-linearly dependent latent variables; an idea well in line with the usual motivations behind deep neural networks such as efficient, compositional representation \cite{lecun_deep_2015}. 
Such hierarchical models are usually formalized by introducing a set of $L$ latent variables  $\zb=\zb_1, \dots, \zb_L$ and letting the generative model be defined as,
%
\begin{equation}
    p_\thetab(\xb|\zb)=p(\xb|\zb_1) p_\thetab(\zb_1|\zb_2) \cdots p_\thetab(\zb_{L-1}|\zb_L) p(\zb_L) \enspace .
\end{equation}
%
This top-down generative model can be efficiently sampled via ancestral sampling; first $\zb_L$ is drawn from the prior $p(\zb_L)$, and then each $\zb_l$ is drawn from the corresponding $p(\zb_l|\zb_{l+1})$ until we can draw $p(\xb|\zb_1)$.

The inference model can then be defined in two ways respectively referred to as \emph{bottom-up} \parencite{burda_importance_2016}
\begin{equation}
    %q_\phib(\zb|\xb)=q_\phib(\zb_1|\xb)q_\phib(\zb_2|\zb_1)\cdots q_\phib(\zb_L|\zb_{L-1})
    q_\phib(\zb|\xb) = q_\phib(\zb_1|\xb)\prod_{i=2}^{L} q_\phib(\zb_i|\zb_{i-1})
\end{equation}
and \emph{top-down} \parencite{sonderby_ladder_2016}
\begin{equation}
    % q_\phib(\zb|\xb)=q_\phib(\zb_{L}|\xb)q_\phib(\zb_{L-1}|\zb_L)\cdots q_\phib(\zb_2|\zb_1),
    q_\phib(\zb|\xb) = q_\phib(\zb_L|\xb)\prod_{i=1}^{L-1} q_\phib(\zb_{i}|\zb_{i+1}) \ .
\end{equation}
A variant of the variational autoencoder that employs a bidirectional inference network has also been proposed \parencite{maaloe_biva_2019}.
Regardless of the choice of inference model, the resulting hierarchical VAE is trained using the ELBO \cref{eq:variational-autoencoder-elbo} and the reparameterization trick.
Recent architectural advances have alleviated the posterior collapse problem and made it possible to train VAEs with many latent variables \parencite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.

@@

% %

>>>>

\lesstodo[inline]{Come up with a suitable example explaining the intuition behind information entropy.}

@@

\section{Out-of-distribution detection} \label{sec:out-of-distribution-detection}
\lesstodo[inline]{Maybe this section should come after the VAE background?}
% 
% OUTLINE: C.f. https://openaccess.thecvf.com/content/CVPR2023W/VAND/papers/Graham_Denoising_Diffusion_Models_for_Out-of-Distribution_Detection_CVPRW_2023_paper.pdf
% - Formalize the problem. https://arxiv.org/pdf/2302.10326.pdf
%   - Assume we have a dataset $\mathcal{D}_{\text{train}}$ of in-distribution data and we would like to build a model that can be used to assess whether a certain data point $\xb$ is sampled from $\mathcal{D}_{\text{train}}$ or not. Given a test point $\xb$, the model outputs a score $s(\xb)$ that indicates whether $\xb$ is in-distribution or out-of-distribution. 
%   - In the context of neural networks, this problem dates back to at least 1994 \parencite{bishop_novelty_1994}. 
%
% - Categorisation into supervised and unsupervised. 
%   - Supervised: Whether p(y|x) should be trusted for a given x. 
%   - Unsupervised: Whether x should be trusted and used at all. 
%
In this section we introduce out-of-distribution (OOD) detection and connect it to the concept of uncertainty. 
Since the related work sections of the papers in \cref{chp:paper-hierarchical,chp:paper-modelagnostic} are relatively short, we also provide a concise review of the literature on out-of-distribution detection.

\moretodo[inline]{Briefly hold the readers hand at actualize the importance of OODD.}

@@

% No OOD, feature
The final category of supervised methods derive the score from a latent representation of the input. 
Some methods define OOD classes. 
\textcite{huang_mos_2021} group the classes of the target variable $y$ and defines an OOD class for each group. Each training example is then the correct target for one group and an OOD example for all other groups; a kind of hierarchical version of the OOD class of simpler baselines based on noise augmentation \parencite{ren_likelihood_2019}. 
\moretodo[inline]{Does the previous sentence belong here or in a different paragraph? It is not really using the latent space.}
Similarly, to represent a virtual OOD class, \textcite{wang_vim_2022} generate an additional logit by first computing the residual of the input's latent space representation against a principal feature space and then converting it to a valid logit by matching its mean over training samples to the average maximum logits.

@@

\subsection{Variational inference} \label{sec:variational-autoencoders}
To model high-dimensional data, we want to use neural networks to represent the generative process. 
This leads to deep latent variable models which were pioneered by \textcite{kingma_autoencoding_2014,rezende_stochastic_2014}. The authors propose to factor the joint distribution and parameterize it using a neural network with parameters $\thetab$ and let $p_\thetab(\xb,\zb) = p_\thetab(\xb|\zb)p(\zb)$ where $p(\zb)$ is a prior.

This choice makes both the marginal and posterior distributions intractable since they involve an integral over all possible values of the latent variable $\zb$. 
This precludes training with commonly used methods such as expectation maximization and exact maximum likelihood estimation. Since Markov Chain Monte Carlo methods are usually too computationally expensive for large neural networks, this leads to variational inference as a scalable alternative \parencite{jordan_introduction_1999}.

@@

\paragraph{Evidence lower bound}
By approximating the intractable posterior distribution $p(\zb|\xb)$ with a variational distribution $q(\zb)$, we can derive a lower bound on the also intractable marginal log-likelihood $\log p_\theta(\xb)$.
%
 \begin{align}
    \log p_\thetab(\xb)
    &= \log \int p_\thetab(\xb,\zb),\textrm{d}\zb \nonumber \\
    &= \log \int q(\zb) \frac{p_\thetab(\xb,\zb)}{q(\zb)}\,\textrm{d}\zb  \label{eq:marginal-likelihood-with-variational-approximation} \\
    &\geq \int q(\zb) \log \left( \frac{p_\thetab(\xb,\zb)}{q(\zb)} \right) \,\textrm{d}\zb \nonumber \\
    &= \mathbb{E}_{q(\zb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q(\zb)} \right] \equiv \mathcal{L}_{}(\xb;\thetab,q) \enspace . \label{eq:latent-variable-model-elbo}
    % &= \log \int p_\thetab(\xb|\zb)p(\zb),\textrm{d}\zb \nonumber \\
    % &= \log \int q(\zb) \frac{p_\thetab(\xb|\zb)p(\zb)}{q(\zb)}\,\textrm{d}\zb \nonumber \\
    % &\geq \int q(\zb) \log \left( \frac{p_\thetab(\xb|\zb)p(\zb)}{q(\zb)} \right) \,\textrm{d}\zb \nonumber \\
    % &\geq \mathbb{E}_{q(\zb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q(\zb)} \right] \equiv \mathcal{L}_{}(\xb;\thetab,\phib) \enspace .
\end{align}
%
Since we can evaluate the generative model $p_\thetab(\xb,\zb)$ and are free to define $q(\zb)$ as we please, this bound can be made tractable.

The usual approach in variational inference is to fit an approximate posterior $q(\zb)$ per data point. However, for large datasets, as is common in natural language processing, this can be too computationally expensive. 
Instead, VAEs amortize the cost of variational inference by parameterizing the approximate posterior with a neural network that transforms any data point $\xb$ into the parameters of the corresponding conditional distribution. We denote this by $q_\phib(\zb|\xb)$ where $\phib$ are the parameters of this inference model. This makes it possible to efficiently infer the approximate posterior distribution for any input $\xb$. 
With amortized variational inference, the ELBO is simply be written as,
% The approximate posterior can also be used to compute a lower bound on the marginal distribution $p_\thetab(\xb)$ via Jensen's inequality, the so-called evidence lower bound (ELBO), 
%
\begin{equation} \label{eq:variational-autoencoder-elbo}
    \mathcal{L}_{}(\xb;\thetab,\phib) \equiv \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb,\zb)}{q_\phib(\zb|\xb)} \right] \leq \log p_\thetab(\xb) \enspace .
\end{equation}
%
Although analytically evaluating the ELBO is intractable due to the expectation, it can be approximated by sampling and then used to train VAEs. We will return to this after examining the ELBO in more detail.

\paragraph{KL-divergence to prior}
Since VAEs factorize the joint distribution, we can also write the ELBO in \cref{eq:variational-autoencoder-elbo} as,
%
\begin{align} \label{eq:variational-autoencoder-elbo-factorized}
    \mathcal{L}_{}(\xb;\thetab,\phib)
    % &\equiv \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
    &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right] + \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p(\zb)}{q_\phib(\zb|\xb)}\right] \nonumber \\
    &= 
    \underset{\text{reconstruction loss}}{\underbrace{
        \mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]
    }}
     - 
    \underset{\text{KL-divergence to prior}}{
        \underbrace{\vphantom{\mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]}{D_\text{KL}\left( q_\phib(\zb|\xb) \parallel p(\zb) \right)}
    }} \enspace .
\end{align}
%
This reveals a commonly used interpretation of the ELBO: 
The expected log-likelihood of the generative model $\mathbb{E}_{q_\phib(\zb|\xb)}\left[\log p_\thetab(\xb|\zb)\right]$ acts as a reconstruction loss that encourages the model to make $q_\phib(\zb|\xb)$ as peaky as possible since mapping many $\zb$ to a single $\xb$ is hard. 
The negative KL-divergence between the approximate posterior $q_\phib(\zb|\xb)$ and the prior $p(\zb)$ acts as a regularizer that forces the approximate posterior to be within the support of the prior (reverse KL-divergence). This helps prevent overfitting but importantly also enables ancestral sampling from the prior to generate high quality data.

@@

We can analyze the inference gap further \cite{cremer_inference_2018}. 
By amortizing the variational inference we introduce a gap compared to the alternative of fitting a posterior per data point. %, $q^{*}_{\xb}(\zb)$.
This is referred to as the \emph{amortization gap} and can be reduced by using large capacity amortized posteriors. 
In case the true posterior does not belong to the chosen class of variational posteriors, the inference gap cannot be reduced to zero even with infinite capacity. The remaining gap is referred to as the \emph{approximation gap}. 
If the inference gap is closed, the ELBO becomes equal to the marginal likelihood, the value of which then depends on how well the generative model is fitted \cite{cremer_inference_2018,frellsen_deep_2019}.

\lesstodo[inline]{Do we want to refer to the parsimony gap?}
\lesstodo[inline]{Maybe add a small graphic illustrating the gaps?}
% A final error is introduced  the maximally attainable likelihood over all the group $G$ of possible models, $\mathcal{l}(G)$ 
% It is important to also note that in fitting a VAE, the approximation gap is not the only source of error.

@@

\paragraph{Sampling}
Since we have freedom to select the approximate posterior, we can choose it such that it is easy to sample and then use that property to form a Monte Carlo estimate of the ELBO, $\mathcal{L}_{}(\xb;\thetab,\phib)$,
% 
\begin{equation} \label{eq:variational-autoencoder-elbo-sampling-expanded}
    % \mathcal{L}_{}(\xb;\thetab,\phib)
    % \approx 
    \widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)
    =
    \frac{1}{S} \sum_{s=1}^{S} \log \frac{p_\thetab(\xb|\zb_{s}) p(\zb_{s})}{q_\phib(\zb_{s}|\xb)} \enspace , 
    % \frac{1}{S} \sum_{s=1}^S \log p_\thetab(\xb|\zb_{s}) + \log p(\zb_{s}) - \log q_\phib(\zb_{s}|\xb) \enspace , 
\end{equation}
%
where $\zb_{s} \sim q_\phib(\zb|\xb)$ are samples from the approximate posterior. 
For some choices of prior and approximate posterior, such as Gaussians, the KL-divergence term of the ELBO can be computed analytically which reduces estimator variance. 
This estimator can be written as,
%
\begin{equation} \label{eq:variational-autoencoder-elbo-sampling-KL}
    \widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)
    \equiv
    \frac{1}{S} \sum_{s=1}^{S} \log p_\thetab(\xb|\zb_{s}) - D_\text{KL}\left( q_\phib(\zb|\xb) \,||\, p(\zb) \right)\enspace .
\end{equation}
%
Since $\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)$ is the sample average, it is an unbiased estimator of the ELBO, $\mathbb{E}\left[\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)\right] = \mathcal{L}_{S}(\xb;\thetab,\phib)$, and hence a biased estimator of the marginal likelihood, $\log p(\xb)$.

Whereas we arrive at $\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)$ by first deriving the ELBO as a lower bound on the marginal likelihood and then estimating that bound by Monte Carlo sampling, \textcite{burda_importance_2016} instead propose to directly Monte Carlo sample the expectation in the logarithm of the intractable marginal likelihood \cref{eq:marginal-likelihood-with-variational-approximation},
%
\begin{equation}
    \widehat{\mathcal{L}}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \equiv \log \left( \frac{1}{S}\sum_{s=1}^{S} \frac{p_\thetab(\xb|\zb_{s})p(\zb_{s})}{q_\phib(\zb_{s}|\xb)} \right) = \frac{1}{S} \sum_{s=1}^S \log w_\thetab(\xb,\zb_{s}) \enspace ,
\end{equation}
%
where $w_\thetab(\xb,\zb_{s})$ are the importance weights. Note the difference with the estimator in \cref{eq:variational-autoencoder-elbo-sampling-expanded}. 
By taking the expectation of this estimator, we can see that it too is a lower bound on the marginal likelihood,
%
\begin{align}
    \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib)
    &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[ \widehat{\mathcal{L}}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \right] \nonumber \\
    &= \mathbb{E}_{\zb_{s}\sim q_\phib(\zb|\xb)}\left[ \frac{1}{S}\sum_{s=1}^{S} \log \left( w_\thetab(\xb,\zb_{s}) \right) \right] \nonumber \\
    &\leq \log \mathbb{E}_{\zb_{s}\sim q_\phib(\zb|\xb)}\left[ \frac{1}{S}\sum_{s=1}^{S} w_\thetab(\xb,\zb_{s}) \right] \enspace . \nonumber \\
    &= \log p(\xb) \enspace .
\end{align}
%
From this and the law of large numbers, it can be shown that for all $S$,
%
\begin{equation}
    \log p(\xb) \geq \mathcal{L}^{\text{IWAE}}_{S+1}(\xb;\thetab,\phib) \geq \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \enspace ,
\end{equation}
%
and $\lim_{S\rightarrow\infty} \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \rightarrow \log p(\xb)$, if $w_\thetab(\xb,\zb_{s})$ are bounded. 
Hence, $\widehat{\mathcal{L}}^{\text{IWAE}}_{S}$ is a biased estimator of $\log p(\xb)$ but with a bias that goes to zero as the number of samples goes to infinity. 
With $S=1$, the estimator is equivalent to the regular ELBO \cref{eq:variational-autoencoder-elbo} and so is always as tight or tighter than the ELBO \parencite{burda_importance_2016}.

\paragraph{Reparameterization} 
% We want to use maximum likelihood estimation to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$. 
% Gradient-based optimization works well for neural networks, but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
Computing the gradient of the ELBO with respect to $\thetab$ is straightforward,
%
\begin{equation}
    % \nabla_\thetab \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] 
    \nabla_\thetab \mathcal{L}_{}(\xb;\thetab,\phib)
    = \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \nabla_\thetab \log p_\thetab(\xb|\zb) \right] \enspace .
\end{equation}
%
However, the gradient with respect to the $\phib$ is more challenging. Since $\phib$ occurs in the expectation itself, we may not move the gradient inside the expectation. 
However, by reparameterizing the latent variable $\zb$ as a deterministic function of the input $\xb$ and a random variable $\epsilonb\sim p(\epsilonb)$, $\zb = g_\phib(\xb,\epsilonb)$, we can write the gradient as a path-wise derivative \parencite{kingma_autoencoding_2014,rezende_stochastic_2014,mohamed_monte_2019},
%
\begin{equation}
    % \nabla_\phib \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] 
    \nabla_\phib \mathcal{L}_{}(\xb;\thetab,\phib)
    = \mathbb{E}_{p(\epsilonb)} \left[ \nabla_\phib \log \frac{p_\thetab\left(\xb|g_\phib(\xb,\epsilonb)\right)p\left(g_\phib(\xb,\epsilonb)\right)}{q_\phib\left(g_\phib(\xb,\epsilonb)|\xb\right)} \right] \enspace . \\
\end{equation}
%
The most important instantiation of this reparameterization is the diagonal Gaussian case. 
By choosing $q_\phib(\zb|\xb) = \mathcal{N}(\zb; \mub_\phib(\xb), \mathbf{I}\sigmab_\phib(\xb))$ it is possible to sample differentiably via the reparameterization trick by setting $g_\phib(\xb,\epsilonb) = \mu_\phib(\xb) + \sigma_\phib(\xb) \odot \epsilonb$ with $\epsilonb\sim\mathcal{N}(0,\mathbf{I})$. 
By assuming the covariance matrix is diagonal, this becomes a mean-field approximation which imposes independence between the dimensions of $\zb$. 
The gradient for $\mathcal{L}^{\text{IWAE}}_{S}$ is derived similarly.

@@

\subsection{Hierarchical models}\label{sec:hierarchical-vae-background}
The mean-field assumption imposed by $\zb \sim\mathcal{N}(\mub,\mathbf{I}\sigmab^2)$ can often be more restrictive for model expressiveness than we would like. 
While learning a full covariance is a simple solution, it is not always sufficiently computationally efficient and, in any case, enables learning only linear covariance between elements of $\zb$. 
The drawbacks of mean-field approximation and linear covariance have lead to research into learning hierarchies of several non-linearly dependent latent variables; an idea well in line with the usual motivations behind deep neural networks such as efficient, compositional representation \cite{lecun_deep_2015}. 
Such hierarchical models are usually formalized by introducing a set of $L$ latent variables  $\zb=\zb^{(1)}, \dots, \zb^{(L)}$ and letting the generative model be defined as,
%
\begin{equation}
    p_\thetab(\xb|\zb)=p(\xb|\zb^{(1)}) p_\thetab(\zb^{(1)}|\zb^{(2)}) \cdots p_\thetab(\zb^{(L-1)}|\zb^{(L)}) p(\zb^{(L)}) \enspace .
\end{equation}
%
This top-down generative model can be efficiently sampled via ancestral sampling; first $\zb^{(L)}$ is drawn from the prior $p(\zb^{(L)})$, and then each $\zb^{(l)}$ is drawn from the corresponding $p(\zb^{(l)}|\zb^{(l+1)})$ until we can draw $p(\xb|\zb^{(1)})$.

The inference model can then be defined in two ways respectively referred to as \emph{bottom-up} \parencite{burda_importance_2016}
\begin{equation}
    %q_\phib(\zb|\xb)=q_\phib(\zb^{(1)}|\xb)q_\phib(\zb^{(2)}|\zb^{(1)})\cdots q_\phib(\zb^{(L)}|\zb^{(L-1)})
    q_\phib(\zb|\xb) = q_\phib(\zb^{(1)}|\xb)\prod_{i=2}^{L} q_\phib(\zb^{(i)}|\zb^{(i-1)})
\end{equation}
and \emph{top-down} \parencite{sonderby_ladder_2016}
\begin{equation}
    % q_\phib(\zb|\xb)=q_\phib(\zb_{L}|\xb)q_\phib(\zb^{(L-1)}|\zb_L)\cdots q_\phib(\zb_2|\zb^{(1)}),
    q_\phib(\zb|\xb) = q_\phib(\zb^{(L)}|\xb)\prod_{i=1}^{L-1} q_\phib(\zb^{(i)}|\zb^{(i+1)}) \ .
\end{equation}
A variant of the variational autoencoder that employs a bidirectional inference network has also been proposed \parencite{maaloe_biva_2019}. 
Regardless of the choice of inference model, the resulting hierarchical VAE is trained using the ELBO \cref{eq:variational-autoencoder-elbo} and the reparameterization trick. 
Recent architectural advances have alleviated the posterior collapse problem and made it possible to train VAEs with many latent variables \parencite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}. A key modelling choice in these works is conditioning each latent variable directly on the input via residual connections.

@@
